Negative Selection in the 20th Century

### 2020-07-15 why-negative-selection.Rmd
Pushing update before reinstalling hard drive

---
title: "Negative Selection in the 20th Century"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "April 2020"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
---

@@

pval_geom_point <- function (size = 1.5, stroke = 0.7, ...) {
  geom_point(shape = 21, size = size, stroke = stroke, ...)
}

@@

add_n <- function (var, famhist_var = var, dv = "n_children") {
  var <- as.factor(var)
  n <- n_in_regs(famhist_var, dv)
  levels(var) <- sprintf("%s (N = %s)", levels(var), 
                         prettyNum(n[levels(var)], big.mark = ","))
  # var_n <- sprintf("%s (N = %s)", var, prettyNum(n[var], big.mark = ","))
  # var_n[is.na(var)] <- NA
  var
}

@@

Existing work suggests that natural selection is taking place in modern populations,
but that effect sizes are small. In this paper we investigate natural selection
in the UK over the twentieth century, using data from UK Biobank. We use
questions on number of children and number of siblings to reconstruct two
generations of family size per respondent. We then relate these to polygenic
scores for several different characteristics.

Natural selection is taking place on most polygenic scores, and is broadly
"negative" in that characteristics that most people would consider undesirable
are being selected for. Effect sizes are larger for number of siblings than for
number of children. This is not because the strength of natural selection has
changed over time. Instead, the difference is probably due to ascertainment bias
in UK Biobank. Indeed, natural selection is concentrated among population
subgroups that are under-represented in the sample, including less-educated
people and poorer households. When we weight the sample to correct for
ascertainment bias, many effect sizes are considerably larger.

Among mothers who were older at first birth, and among parents with fewer sexual
partners, effect signs are reversed. The effects of PGS can be decomposed into
two channels: an effect on age at first live birth, and an opposite-signed
effect on number of children conditional on age at first live birth. This
pattern may be explained by economic theories of fertility. In these, higher
potential earnings have two opposite effects on fertility: a
fertility-increasing income effect (higher income makes children more
affordable), and a fertility-lowering substitution effect (time spent on
childrearing has a higher cost in foregone earnings). Among couples who can take
advantage of coordination in childcare activities, the income effect dominates;
among single parents, or those in unstable relationships, the substitution effect
dominates.

```
# Open questions/TODO

* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1. 
* Reweight other regressions (e.g. the omnibus regression in the appendix)
* Look at geography, esp. in siblings regressions.
* control for age in n_partners regressions.
* can we weight men by age at first child, using GHS? If not, may need
  to plot the age_flb weighted regressions separately, against unweighted
  women-only regressions.
* Can we find representative data for weights by n_partners?
```

@@

drake::loadd(famhist)
drake::loadd(ghs_weights)
famhist <- left_join(famhist, ghs_weights, by = "f.eid")
pgs_over_time <- famhist %>% 
      group_by(
        YOB = santoku::chop(YOB, 
          seq(1940, 1970, 5), 
          labels    = seq(40, 65, 5),
          extend    = FALSE
        ) 
      ) %>% 
      filter(! is.na(YOB)) %>% 
      summarize(across(all_of(score_names), ggplot2::mean_cl_normal, na.rm = TRUE)) %>% 
      pivot_longer(-YOB, names_to = "score_name", values_to = "score") %>% 
      mutate(
        score_lo = score$ymin,
        score_hi = score$ymax,
        score    = score$y
      )

@@

Figures \@ref(fig:plot-educ-level) and \@ref(fig:plot-income) plot effect sizes
for each polygenic score among different subgroups of the sample. Figure
\@ref(fig:plot-educ-level) groups respondents by age of completing full-time
education. Figure \@ref(fig:plot-income) groups respondents by household income.
Effect sizes are larger and more significant for the lowest income category, and
for the lowest education category. Note that the overall effect is not a simple
average of the effect among the different subgroups, because polygenic scores 
may also shift respondents between the subgroups. For example, a high 
PGS for educational attainment will increase the age at which a respondent 
leaves school.

@@

Figure \@ref(fig:plot-sexes) shows effect sizes of PGS on number of children
separately for males and females. Differences are particularly large for
educational attainment, height and MDD. There is no overall pattern in
differences of effect size, however.

@@

drake::loadd(res_wt_ghs_weights)
drake::loadd(res_wt_flb_weights)
drake::loadd(res_wt_msoa_weights)
drake::loadd(res_wt_age_qual_weights)
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_all)
drake::loadd(res_sex)

@@

# TODO: move later
res_weighted_all <- bind_rows(
        "N children" = res_wt_ghs_weights, 
        "N siblings" = res_sibs_parent_weights, 
      .id = "dep.var") 
res_weighted_all$reg.type <- "weighted"
res_weighted_all <- bind_rows(
        res_weighted_all,
        res_all %>% filter(reg.type == "raw") %>% select(-score_name)
      )

res_wide <- full_join(
        res_wt_ghs_weights,
        res_sibs_parent_weights,
        by = "term",
        suffix = c(".children", ".sibs")
      ) %>% 
      filter(p.value.children < 0.05/33 | p.value.sibs < 0.05/33)

chn_larger <- with(res_wide, estimate.children > estimate.sibs)

@@

Effect sizes are reasonably consistent across the geographical and
age/qualification weights. But they are increased again when we weight by age at
first live birth.This suggests that estimates might be increased still more by
weighting on other relevant demographic variables. Overall, these results
suggest that previous work may have underestimated the level of selection
occurring in the population. [TODO: who?]

@@

To compensate for this, we reweight our preferred weightings by the inverse of
*number of siblings*.

TODO: 
- weight regressions using "best effort" weights
- look at change over time to confirm your "no change" hypothesis
- look at siblings, again using "best" weights

We rerun the basic regressions for both children and siblings, using the new
weights. Figure \@ref(fig:plot-res-weighted) shows the results. The weightings
are fairly successful at reconciling the overall size of effects from siblings
and children regressions. Among weighted scores that are significant (at p <
0.05/33) in either regression, `r sum(chn_larger)` estimates are
higher for children and `r length(chn_larger) - sum(chn_larger)` are
lower. However, some estimates differ widely or change
sign across the generations.

```{r plot-res-weighted, fig.cap = "Effects of polygenic scores on number of children/siblings. Regressions weighted by sex, age and education."}

res_weighted_all %>% 
      rename(
        "Dependent variable" = dep.var, 
      ) %>% 
      filter(reg.type == "weighted") %>% 
      standard_ggplot(fill = `Dependent variable`, n_regs = 33, order_idx = 1, 
        shape = reg.type) + 
        scale_shape_manual(values = c(21, 4))

@@

Why do selection effects have *opposite* signs across certain subgroups -- for
example, across older and younger mothers by age at first live birth, or across those
with more and fewer lifetime sexual partners? A possible answer is given by the
economic theory of fertility [TODO cite Becker 1960]. According to this,
increases in potential earned income affect fertility via two opposing channels.
There is an "income effect" by which children become more affordable, like any
other good. Since childrearing has a cost in time, there is also a substitution
effect: the opportunity cost of childrearing increases if one's market wage is
higher. The income effect would lead higher earners to have more children. The
substitution effect would lead higher earners to have fewer children. It is
often assumed that the substitution effect will dominate for lone parents, since
they have less opportunity to share childcare responsibilities, while the income
effect will dominate for couples who are able to reap the gains from
specialization.

@@

# Bits to integrate

We can run similar regressions for the parents' generation, using the subsets of respondents who reported their mother's or father's age
and who had no elder siblings. We run *sibling* regressions on these subsets,
controlling for either parent's age at their birth.
Figure \@ref(fig:plot-age-birth-parents) shows
the results. Effect sizes are very similar, whether controlling for
father's or mother's age at respondent's birth or mother's age at respondent's birth. Unlike for the respondents' generation, effect sizes are positively correlated with
the effect sizes from bivariate regressions (father's age at birth: 
$\rho$ `r cor_raw_fath`; mother's age at birth: $\rho$ `r cor_raw_moth`).

These results suggest that polygenic scores may directly correlate with
age at first live birth. Figure \@ref(fig:plot-age-flb-dv) plots
estimated effect sizes from bivariate regressions for respondents, and
Figure \@ref(fig:plot-age-birth-parents-dv) does the same for their parents. 
Effect sizes are reasonably large. They are also very highly correlated across 
generations. Effect sizes of PGS on father's age at own birth, and on own age
at first live birth, have a correlation of `r cor_flb_fath`; for mother's age
and own age it is `r cor_flb_moth`

@@

drake::loadd(res_period)
res_period %<>% 
      tidyr::extract(term, c("period", "score_name"),
        regex="year_split(.*):(.*)")
n_tests <- nrow(res_period)/2

@@

To check whether effect sizes were changing over time, we ran regressions 
interacting PGS with birth year, median split at 1950. 
Tables \@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)
summarize the results. Very few scores are significantly different. Only
one score, EA3, changes significantly across time in both generations: the
absolute size of the (negative) effect significantly increases in sibling 
regressions, and significantly decreases in children regressions.

@@

summary_period %>% 
      group_by(Change_Parents) %>% 
      count() %>% 
      period_table("Change in effect sizes between early and late born parents, 'sibling' regressions")

@@

```{r tbl-children-by-period}
summary_period %>% 
      group_by(Change_Respondents) %>% 
      count() %>% 
      period_table("Change in effect sizes between early and late born respondents, 'children' regressions")
```

@@

>>>>

---
title: "Negative Selection in the 20th Century"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "April 2020"
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
---

@@

pval_geom_point <- function (size = 1.5, stroke = 0.5, ...) {
  geom_point(shape = 21, size = size, stroke = stroke, ...)
}

@@

add_n <- function (var, famhist_var = var, dv = "n_children") {
  var <- as.factor(var)
  n <- n_in_regs(famhist_var, dv)
  levels(var) <- sprintf("%s (N = %s)", levels(var), 
                         prettyNum(n[seq_len(nlevels(var))], big.mark = ","))
  # var_n <- sprintf("%s (N = %s)", var, prettyNum(n[var], big.mark = ","))
  # var_n[is.na(var)] <- NA
  var
}

@@

Existing work suggests that natural selection is taking place in modern populations,
but that effect sizes are small. In this paper we investigate natural selection
in the UK over the twentieth century, using data from UK Biobank. We use
questions on number of children and number of siblings to reconstruct two
generations of family size per respondent. We then relate these to 33 different
polygenic scores for a variety of  characteristics.

Natural selection is taking place on most polygenic scores, and is concentrated
among population subgroups that are under-represented in the sample, including
less-educated people and poorer households. When we weight the sample to correct
for ascertainment bias, many effect sizes are considerably larger.

Among mothers who were older at first birth, and among parents with fewer sexual
partners, effect signs are reversed. The effects of PGS can be decomposed into
two channels: an effect on age at first live birth, and an opposite-signed
effect on number of children conditional on age at first live birth. This
pattern can be explained by economic theories of fertility. In these, higher
potential earnings have two opposite effects on fertility: a
fertility-increasing income effect (higher income makes children more
affordable), and a fertility-lowering substitution effect (time spent on
childrearing has a higher cost in foregone earnings). Among couples who can take
advantage of coordination in childcare activities, the income effect dominates;
among single parents, or those in unstable relationships, the substitution effect
dominates.

```{r, eval = FALSE}
# Open questions/TODO

* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1. 
* Reweight other regressions (e.g. the omnibus regression in the appendix)
* Look at geography, esp. in siblings regressions.
* control for age in n_partners regressions.
* Beauchamp 2016 used the HRS. He reckons it's representative. Is it? How do
  his effect sizes compare?
  - he also worries about the quality/quantity tradeoff, which might mean that
    people with fewer children have more grandchildren (than you'd expect). 
    As we have the two generations, we can test this. (Actually, there's a 
    positive correlation of about 0.1)

@@

drake::loadd(famhist)

pgs_over_time <- famhist %>% 
      group_by(
        YOB = santoku::chop(YOB, 
          seq(1940, 1970, 5), 
          labels    = seq(40, 65, 5),
          extend    = FALSE
        ) 
      ) %>% 
      filter(! is.na(YOB)) %>% 
      summarize(across(all_of(score_names), ggplot2::mean_cl_normal, na.rm = TRUE)) %>% 
      pivot_longer(-YOB, names_to = "score_name", values_to = "score") %>% 
      mutate(
        score_lo = score$ymin,
        score_hi = score$ymax,
        score    = score$y
      )

@@

Figures \@ref(fig:plot-educ-level) and \@ref(fig:plot-income) plot effect sizes
for each polygenic score among different subgroups of the sample. Figure
\@ref(fig:plot-educ-level) groups respondents by age of completing full-time
education. Figure \@ref(fig:plot-income) groups respondents by household income.
Effect sizes are larger and more significant for the lowest income category, and
for the lowest education category. Note that the overall effect is not a simple
average of the effect among the different subgroups, because polygenic scores 
may also shift respondents between the subgroups. For example, a high 
PGS for educational attainment may increase the age at which a respondent 
leaves school.

@@

Figure \@ref(fig:plot-sexes) shows effect sizes of PGS on number of children
separately for males and females. Differences are particularly large for
educational attainment, height and MDD. Several PGS for mental illness and
personality traits are more selected for (or less against) among women, including 
major depressive disorder (MDD), schizophrenia and neuroticism, while 
extraversion is more selected for among men. Waist circumference and waist-hip 
ratio are less selected for among women, and educational attainment is more
selected against, though the difference is only significant for EA3.

@@

drake::loadd(res_wt_flb_weights)
drake::loadd(res_wt_msoa_weights)
drake::loadd(res_wt_age_qual_weights)
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_all)
drake::loadd(res_sex)

@@

@@

Effect sizes are reasonably consistent across the geographical and
age/qualification weights, but they are increased again when we weight by age at
first live birth. This suggests that estimates might be further affected by
weighting on other demographic variables. Given that the Biobank sample
seems to be ascertained in ways that reduce estimates of natural selection, 
we view our estimates as likely to be conservative.

@@

To compensate for this, we reweight our preferred weightings (Age/Qualification)
by the inverse of *number of siblings*. We then regress this on PGS.
Figure \@ref(fig:plot-res-sibs) shows the results. For comparison, we show
regressions on *number of children*, where we have excluded respondents with no 
children.

Effect sizes appear larger in the parents' generation, i.e.
for *number of siblings*. However, we treat this result cautiously, because
when we split respondents up by year of birth, we find few differences in
effect sizes between early- and late-born respondents, whether our 
dependent variable is *number of siblings* or *number of children*. In other words,
since estimated effect sizes change when we change the dependent variable,
but do not change over time within either dependent variable, the explanation
may be due to remaining ascertainment bias within our sample.

```{r plot-res-sibs, fig.cap = "Effects of polygenic scores on fertility, respondents' parents vs. respondents"}
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_children_comparison)

res_sibs_kids <-  bind_rows(
                    Siblings = res_sibs_parent_weights, 
                    Children = res_children_comparison,
                    .id = "Dep. var."
                  )

standard_ggplot(res_sibs_kids, `Dep. var.`, n_regs = length(score_names))
```

@@

Why do selection effects have *opposite* signs across certain subgroups -- for
example, across older and younger mothers by age at first live birth, or across those
with more and fewer lifetime sexual partners? A possible answer is given by the
economic theory of fertility [@becker1960economic]. According to this,
increases in potential earned income affect fertility via two opposing channels.
There is an "income effect" by which children become more affordable, like any
other good. Since childrearing has a cost in time, there is also a substitution
effect: the opportunity cost of childrearing increases if one's market wage is
higher. The income effect would lead higher earners to have more children. The
substitution effect would lead higher earners to have fewer children. It is
often assumed that the substitution effect will dominate for lone parents, since
they have less opportunity to share childcare responsibilities, while the income
effect will dominate for couples who are able to reap the gains from
specialization.

@@

# Conclusion

Natural selection is taking place in modern populations on variants underlying
many polygenic scores. In a loose sense, phenotypes relating to scores which are
selected for (against) would be seen as undesirable (desirable) by many people.
For example, most people would probably prefer to have high educational
attainment, a low risk of major depressive disorder, and a low risk of coronary
artery disease, but PGS underlying these traits are being selected for.
Potentially, natural selection could increase the health burden on modern
societies, but that depends on effect sizes. Existing work argues that these are
small. However, as we show, estimates can be crucially affected by ascertainment
bias in population samples. In addition, we do not know whether estimated effect
sizes of natural selection will increase or decrease as more accurate polygenic
scores are produced. We are also unsure about whether polygenic scores for new
phenotypes will show a similar pattern of natural selection as existing ones. In
short, it is probably too early to tell whether modern natural selection is
substantively important or not.

Previous work has documented natural selection in modern populations
[@beauchamp2016genetic; @kong2017selection; @sanjak2018evidence] but has not
explained its underlying causes. The economic theory of fertility provides a
parsimonious explanation for which traits are selected, and for why the
direction of selection is reversed among certain subgroups. An alternative
theory is that traits selected for are linked to impulsivity and low time
discount rates, via the channel of early sexual behaviour. The data here provide
some support for this: scores which might plausibly be linked to
impulsivity,like ADHD and age at smoking initiation, are selected for. However,
the "impulsivity" theory is less good at explaining variation in selection
across the full range of scores, including e.g. waist-hip ratio and BMI.
Impulsivity also does not explain the reversed sign of selection among e.g.
older parents. Overall, then, the economic theory is probably the leading
explanation at present.

@@

drake::loadd(res_period_parents)
drake::loadd(res_period_children)

res_period <- bind_rows(res_period_parents, res_period_children)

res_period %<>% 
      tidyr::extract(term, c("period", "score_name"),
        regex="year_split(.*):(.*)")
n_tests <- nrow(res_period)/2

@@

To check whether effect sizes were changing over time, we ran regressions 
interacting PGS with birth year, median split at 1950 ("early born" versus "late born"). We use both 
*number of children* and *number of siblings* as a dependent variable. We weight 
using age/qualification cells, and further adjust for selection in the parents'
generation (see above).

Tables \@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)
summarize the results. There is no change in natural selection in the parents'
generation. In the respondents' generation, effect sizes were larger in
absolute size among the late born for five PGS: autism, cognitive ability,
EA2, EA3 and extraversion. Three PGS showed significant changes in sign: 
alcohol use (positive to negative); conscientiousness (negative to positive);
and type II diabetes (negative to positive).

@@

summary_period %>% 
      group_by(Change_Parents) %>% 
      count() %>% 
      period_table("Change in effect sizes between parents of early and late born respondents (regressions on number of siblings)")

@@

```{r tbl-children-by-period}
summary_period %>% 
      group_by(Change_Respondents) %>% 
      count() %>% 
      period_table("Change in effect sizes between early and late born respondents (regressions on number of children)")
```

## Selection controlling for age at first live birth: respondents' parents

Among the parents' generation, we can control for age at first live birth using
the subsets of respondents who reported their mother's or father's age, and who
had no elder siblings. We run regressions on *number of siblings* on these
subsets, controlling for either parent's age at their birth. Figure
\@ref(fig:plot-age-birth-parents) shows the results. Effect sizes are very
similar, whether controlling for father's or mother's age. Unlike for the
respondents' generation, effect sizes are positively correlated with the effect
sizes from bivariate regressions without the age at birth control (father's age
at birth: $\rho$ `r cor_raw_fath`; mother's age at birth: $\rho$ `r
cor_raw_moth`).

## Effects of PGS on age at first live birth

Our results suggest that polygenic scores may directly correlate with
age at first live birth. Figure \@ref(fig:plot-age-flb-dv) plots
estimated effect sizes from bivariate regressions for respondents, and
Figure \@ref(fig:plot-age-birth-parents-dv) does the same for their parents. 
Effect sizes are reasonably large. They are also very highly correlated across 
generations. Effect sizes of PGS on father's age at own birth, and on own age
at first live birth, have a correlation of `r cor_flb_fath`; for mother's age
and own age it is `r cor_flb_moth`

@@

# References
